RProtoBuf: Efficient Cross-Language Data Serialization in R

### 2013-12-21 papers/rjournal/eddelbuettel-francois-stokely.Rnw
Flesh out the "Under the hood: S4 Classes, Methods, and Pseudo Methods" section by describing how we wrap over 100 functions of 6 main classes with Rcpp.  How this exercise partially motivated the development of Rcpp Modules, and add concise tables of the methods for descriptors and field descriptors rather than going through and providing examples of every function like we do in the other vignette.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@589 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The \CRANpkg{RProtoBuf} package uses the S4 system to store
information about descriptors and messages, but the information stored
in the R object is very minimal and mainly consists of an external
pointer to a C++ variable that is managed by the \texttt{protobuf} C++
library.

Using the S4 system allows the \texttt{RProtoBuf} package to dispatch
methods that are not generic in the S3 sense, such as \texttt{new} and
\texttt{serialize}.

@@

TODO(ms): Perhaps a table here of the different S4 classes, how many
methods they include, whether it dynamically does dispatch on other
strings, whether/how it is available in the search path, etc.

@@

The \texttt{Message} S4 class represents Protocol Buffer Messages and
is the core abstraction of \CRANpkg{RProtoBuf}.  Each \texttt{Message}
has a \texttt{Descriptor} S4 class which defines the schema of the
data defined in the Message, as well as a number of
\texttt{FieldDescriptors} for the individual fields of the message.

represented in R using the \texttt{Message}
S4 class. The class contains the slots \texttt{pointer} and \texttt{type} as
described on the Table~\ref{Message-class-table}.

@@

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{slot} & \textbf{description} \\
\hline
\texttt{pointer} & external pointer to the \texttt{Descriptor} object of the C++ proto library. Documentation for the
\texttt{Descriptor} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.descriptor.html#Descriptor} \\
\hline
\texttt{type} & fully qualified path of the message type. \\
\hline
\end{tabular}
\caption{\label{Descriptor-class-table}Description of slots for the \texttt{Descriptor} S4 class}
\end{table}

Similarly to messages, the \verb|$| operator can be used to extract
information from the descriptor, or invoke pseuso-methods.

\subsubsection{Extracting descriptors}

The \verb|$| operator, when used on a descriptor object retrieves
descriptors that are contained in the descriptor.

This can be a field descriptor (see section~\ref{subsec-field-descriptor} ),
an enum descriptor (see section~\ref{subsec-enum-descriptor}) or a descriptor
for a nested type

@@

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{new} & Creates a prototype of a message described by this descriptor.\\
\texttt{read} & Reads a message from a file or binary connection.\\
\texttt{readASCII} & Read a message in ASCII format from a file or
text connection.\\
\hline
\texttt{name} & Retrieve the name of the message type associated with
this descriptor.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}) \\
\hline
\texttt{fileDescriptor} & Retrieve the file descriptor of this
descriptor.\\
\texttt{containing\_type} & Retrieve the descriptor describing the message type containing this descriptor.\\
\texttt{field\_count} & Return the number of fields in this descriptor.\\
\texttt{field} & Return the descriptor for the specified field in this descriptor.\\
\texttt{nested\_type\_count} & The number of nested types in this descriptor.\\
\texttt{nested\_type} & Return the descriptor for the specified nested
type in this descriptor.\\
\texttt{enum\_type\_count} & The number of enum types in this descriptor.\\
\texttt{enum\_type} & Return the descriptor for the specified enum
type in this descriptor.\\
\hline
\end{tabular}
\end{small}
\caption{\label{Descriptor-methods-table}Description of methods for the \texttt{Descriptor} S4 class}
\end{table}

>>>>

The \CRANpkg{RProtoBuf} package uses the S4 system to store
information about descriptors and messages.  Using the S4 system
allows the \texttt{RProtoBuf} package to dispatch methods that are not
generic in the S3 sense, such as \texttt{new} and
\texttt{serialize}.

Each R object stores an external pointer to an object managed by
the \texttt{protobuf} C++ library.
The \CRANpkg{Rcpp} \citep{eddelbuettel2011rcpp} package is used to
facilitate the integration of the R and C++ code for these objects.

% Message, Descriptor, FieldDescriptor, EnumDescriptor,
% FileDescriptor, EnumValueDescriptor
%
% grep RPB_FUNC * | grep -v define|wc -l
% 84
% grep RPB_ * | grep -v RPB_FUNCTION | grep METHOD|wc -l
% 33

There are over 100 C++ functions that provide the glue code between
the member functions of the 6 primary Message and Descriptor classes
in the protobuf library.  Wrapping each method individually allows us
to add user friendly custom error handling, type coercion, and
performance improvements at the cost of a more verbose
implementation.  The RProtoBuf implementation in many ways motivated
the development of Rcpp Modules \citep{eddelbuettel2010exposing},
which provide a more concise way of wrapping C++ functions and classes
in a single entity.

@@

\emph{TODO(ms): Perhaps a table here of the different S4 classes, how many
methods they include, whether it dynamically does dispatch on other
strings, whether/how it is available in the search path, etc.}

@@

The \texttt{Message} S4 class represents Protocol Buffer Messages and
is the core abstraction of \CRANpkg{RProtoBuf}.  The class contains
the slots \texttt{pointer} and \texttt{type} as described on the
Table~\ref{Message-class-table}.

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{Message} object of the C++ proto library. Documentation for the
\texttt{Message} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.message.html#Message} \\
\hline
\texttt{type} & Fully qualified name of the message. For example a \texttt{Person} message
has its \texttt{type} slot set to \texttt{tutorial.Person} \\
\hline
\end{tabular}
\caption{\label{Message-class-table}Description of slots for the \texttt{Message} S4 class}
\end{table}

Each \texttt{Message} contains a pointer to a \texttt{Descriptor}
which defines the schema of the data defined in the Message, as well
as a number of \texttt{FieldDescriptors} for the individual fields of
the message.  In addition to the field name extractors of
\texttt{Messages} introduced in the previous section, a complete list
of Message methods is available in Table~\ref{Message-methods-table}.

@@

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{Descriptor} object of the C++ proto library. Documentation for the
\texttt{Descriptor} class is available from the protocol buffer project page:
\url{http://code.google.com/apis/protocolbuffers/docs/reference/cpp/google.protobuf.descriptor.html#Descriptor} \\
\hline
\texttt{type} & Fully qualified path of the message type. \\
\hline
\end{tabular}
\caption{\label{Descriptor-class-table}Description of slots for the \texttt{Descriptor} S4 class}
\end{table}

Similarly to messages, the \verb|$| operator can be used to retrieve
descriptors that are contained in the descriptor, or invoke
pseudo-methods.  Thise can be used to extract field descriptors, enum
descriptors, or descriptors for a nested type.

@@

Table~\ref{Descriptor-methods-table} provides a complete list of the
avalailable methods for Descriptors.

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{new} & Creates a prototype of a message described by this descriptor.\\
\texttt{read} & Reads a message from a file or binary connection.\\
\texttt{readASCII} & Read a message in ASCII format from a file or
text connection.\\
\hline
\texttt{name} & Retrieve the name of the message type associated with
this descriptor.\\
\texttt{as.character} & character representation of a descriptor\\
\texttt{toString} & character representation of a descriptor (same as \texttt{as.character}) \\
\texttt{as.list} & return a named
list of the field, enum, and nested descriptors included in this descriptor.\\
\texttt{asMessage} & return DescriptorProto message. \\
\hline
\texttt{fileDescriptor} & Retrieve the file descriptor of this
descriptor.\\
\texttt{containing\_type} & Retrieve the descriptor describing the message type containing this descriptor.\\
\texttt{field\_count} & Return the number of fields in this descriptor.\\
\texttt{field} & Return the descriptor for the specified field in this descriptor.\\
\texttt{nested\_type\_count} & The number of nested types in this descriptor.\\
\texttt{nested\_type} & Return the descriptor for the specified nested
type in this descriptor.\\
\texttt{enum\_type\_count} & The number of enum types in this descriptor.\\
\texttt{enum\_type} & Return the descriptor for the specified enum
type in this descriptor.\\
\hline
\end{tabular}
\end{small}
\caption{\label{Descriptor-methods-table}Description of methods for the \texttt{Descriptor} S4 class}
\end{table}

\subsection{field descriptors}
\label{subsec-field-descriptor}

The class \emph{FieldDescriptor} represents field
descriptor in R. This is a wrapper S4 class around the
\texttt{google::protobuf::FieldDescriptor} C++ class.
Table~\ref{fielddescriptor-methods-table} describes the methods
defined for the \texttt{FieldDescriptor} class.

\begin{table}[h]
\centering
\begin{tabular}{|cp{10cm}|}
\hline
\textbf{Slot} & \textbf{Description} \\
\hline
\texttt{pointer} & External pointer to the \texttt{FieldDescriptor} C++ variable \\
\hline
\texttt{name} & Simple name of the field \\
\hline
\texttt{full\_name} & Fully qualified name of the field \\
\hline
\texttt{type} & Name of the message type where the field is declared \\
\hline
\end{tabular}
\caption{\label{FieldDescriptor-class-table}Description of slots for the \texttt{FieldDescriptor} S4 class}
\end{table}

\begin{table}[h]
\centering
\begin{small}
\begin{tabular}{l|l}
\hline
\textbf{Method} & \textbf{Description} \\
\hline
\hline
\texttt{as.character} & Character representation of a descriptor\\
\texttt{toString} & Character
representation of a descriptor (same as \texttt{as.character}) \\
\texttt{asMessage} & Return FieldDescriptorProto message. \\
\texttt{name} & Return the name of the field descriptor.\\
\texttt{fileDescriptor} & Return the fileDescriptor where this field is defined.\\
\texttt{containing\_type} & Return the containing descriptor of this field.\\
\texttt{is\_extension} & Return TRUE if this field is an extension.\\
\texttt{number} & Gets the declared tag number of the field.\\
\texttt{type} & Gets the type of the field.\\
\texttt{cpp\_type} & Gets the C++ type of the field.\\
\texttt{label} & Gets the label of a field (optional, required, or repeated).\\
\texttt{is\_repeated} & Return TRUE if this field is repeated.\\
\texttt{is\_required} & Return TRUE if this field is required.\\
\texttt{is\_optional} & Return TRUE if this field is optional.\\
\texttt{has\_default\_value} & Return TRUE if this field has a default value.\\
\texttt{default\_value} & Return the default value.\\
\texttt{message\_type} & Return the message type if this is a message type field.\\
\texttt{enum\_type} & Return the enum type if this is an enum type field.\\
\hline
\end{tabular}
\end{small}
\caption{\label{fielddescriptor-methods-table}Description of methods for the \texttt{FieldDescriptor} S4 class}
\end{table}
